Verifying Spatial Properties of Array Computations

### 2017-04-16 oopsla17/algorithms.tex
start of putting in analysis and inference

\documentclass[10pt,preprint,numbers]{sigplanconf}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}
\usepackage{enumitem}
%\usepackage{subcaption}
\usepackage{siunitx} % For pretty-printing numeric values and SI units
                     % of measure. e.g., the tabular column type S is
                     % used to print nice-looking tables of numbers.
\sisetup{ % defaults
  group-separator={,},
  group-minimum-digits={3},
  output-decimal-marker={.},
  table-format = 6
}

\usepackage{tikz}
\usepackage{xypic}
\usepackage{natbib}

\usetikzlibrary{calc}

%% http://tex.stackexchange.com/questions/55068/is-there-a-tikz-equivalent-to-the-pstricks-ncbar-command
\tikzset{
    ncbar angle/.initial=90,
    ncbar/.style={
        to path=(\tikztostart)
        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
        -- (\tikztotarget)
    },
    ncbar/.default=0.5cm,
}

\tikzset{round left paren/.style={ncbar=0.4cm,out=110,in=-110}}
\tikzset{round right paren/.style={ncbar=0.4cm,out=70,in=-70}}

\fvset{
  linenos=true,
  fontsize=\footnotesize,
  breaklines=true,
 breakafter=w),
  xleftmargin=\parindent,
  numbersep=1em
}

\usemintedstyle{vs}

\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}

\usepackage{hyperref}
\hypersetup{
    unicode=false,          % non-Latin characters in Acrobat's bookmarks
    pdftoolbar=true,        % show Acrobat toolbar?
    pdfmenubar=true,        % show Acrobat menu?
    pdffitwindow=false,      % page fit to window when opened
    pdftitle={},    % title
    pdfauthor={}
    pdfsubject={},   % subject of the document
    pdfnewwindow=true,      % links in new window
    pdfkeywords={keywords}, % list of keywords
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=darkblue,          % color of internal links
    citecolor=darkblue,        % color of links to bibliography
    filecolor=green,      % color of file links
    urlcolor=blue,          % color of external links
}

\usepackage{cleveref}
\crefname{section}{\S}{\S\S}
\Crefname{section}{\S}{\S\S}
\crefname{subsection}{\S}{\S\S\S}
\Crefname{subsection}{\S}{\S\S\S}
\crefname{figure}{fig.}{}
\Crefname{figure}{Fig.}{}
\crefname{definition}{def.}{}
\Crefname{definition}{Def.}{}

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}

\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
\definecolor{orange}{rgb}{0.8,0.4,0.0}
\definecolor{darkorange}{rgb}{0.5,0.2,0.0}
\definecolor{marco}{rgb}{0.0,0.3,0.5}
\definecolor{gray}{rgb}{0.2,0.2,0.2}

\newcommand\mdef{\stackrel{\mathclap{\normalfont\mbox{def}}}{=}}

\newcommand{\bn}{\mathbb{N}}

\newcommand{\todo}[1]{\textcolor{blue}{Dom: #1}}
\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}

\newcounter{block}

\newtheorem{lemma}[block]{Lemma}
\newtheorem{proposition}[block]{Proposition}

\theoremstyle{definition}

\newtheorem{theorem}[block]{Theorem}
\newtheorem{remark}[block]{Remark}
\newtheorem{example}[block]{Example}
\newtheorem{definition}[block]{Definition}

% Writing macros
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}

\newcommand{\dimId}{\texttt{dim}}

% Semantics related
\newcommand{\interp}[1]{\llbracket{#1}\rrbracket}

% Syntax macros
\newcommand{\nonterm}[1]{\textit{#1}}
\newcommand{\term}[1]{\texttt{#1}}

\newcommand{\stenRefl}[1]{\term{pointed} \, (\term{dim=}#1)}
\newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\stenBwd}[3]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\stenCen}[3]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\irrefl}{\texttt{nonpointed}}

\newcommand{\stenReflS}[1]{\term{pointed} \, (\term{dim=}#1)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\irreflS}{\texttt{np}}

\newcommand{\stenFwdSR}[3]{\term{fwd} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenBwdSR}[3]{\term{bwd} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenCenSR}[3]{\term{cen} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenReflSR}[1]{\term{point} (\term{dim=}#1)}

% SYNTAX OPERATIONS AND PREDICATES
%\newcommand{\neigh}{\textsf{neigh}}
\newcommand{\arrayTy}{\textsf{array}}
\newcommand{\rhsExp}{\textsf{rhsExp}}
\newcommand{\var}{\textsf{var}}

%% VECTOR NOTATIONS
\newcommand{\vect}[1]{\textbf{#1}}

\newcommand{\vtwohs}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{cc}#1 \, & \, #2\end{array}\right]\!}

\newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}

\newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{ccc}#1 \; & \; #2 \; & \; #3\end{array}\right]}
\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}#1\\#2\end{array}\right]}
\newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}$#1$\\$#2$\\$#3$\end{array}\right]}
\newcommand{\stwo}[4]
%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
{\setlength{\arraycolsep}{0.1em}
\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}

\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}

%% OPERATIONS ON SPANS and VECTORS
\newcommand{\containedin}{\sqsubseteq}

%% MODEL
\newcommand{\effdims}[2]{\mathit{constr}(#1)_{#2}}

\newcommand{\trule}[1]{{\footnotesize{(\text{#1})}}}
%\include{results}

%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
\newcommand{\consName}{\textbf{agree\textprime}}
\newcommand{\cons}[2]{#1 \,\, \consName{} \,\, #2}
\newcommand{\consAName}{\textbf{agree}}
\newcommand{\consA}[2]{#1 \,\, \consAName{} \,\, #2}
\newcommand{\consSub}[3]{#2 \,\, \consAName{}_{#1} \,\,#3}

\title{Verifying Spatial Properties of Stencil Computations}
%\titlebanner{DRAFT -- Do not distribute}
\authorinfo{}{}{}
%\preprintfooter{DO NOT DISTRIBUTE}

\begin{document}
\maketitle

\bibliographystyle{plainnat}

\begin{abstract}
Stencil computations are a common programming pattern in numerical
  code and typically form the core of graphics and scientific
  computing applications. However, they are often complex and
  unwieldy, involving substantial fine-grained manipulation of array
  indices across multiple arrays and dimensions. This requires careful
  testing, which is almost always done manually and then discarded.
  In this paper, we propose an alternate approach:
  a specification language and automated tool for verifying the
  spatial behaviour of stencils. The specification language is based
  on the hypothesis that stencil computations read from arrays with a
  regular, fixed pattern which can be captured by a simple
  set of abstract combinators. These
  specifications can be embedded as annotations in source code,
  against which code can be checked for conformance. For legacy code,
  spatial specifications can be inferred, providing documentation and
  aiding future software maintenance. We evaluate our specification
  language and verification tool against
  a corpus of numerical Fortran code ($\sim$ 1 million lines) for
  which we generate $\sim$60,000 specifications, showing the vast majority
  of stencil computations indeed have a simple, regular, static shape. % We introduce the
  %language, checking and inference procedures, and its evaluation.
\end{abstract}

%\keywords{program verification, specification, stencils}

\section{Introduction}\label{sec:intro}

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array whose value at each index $i$ is
calculated from a \emph{neighbourhood} of values around $i$ in some
input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following computes the one-dimensional discrete Laplace transform (an
approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
end do
\end{minted}
%
Values \texttt{b(i)} are calculated from a neighbourhood of elements
around \texttt{i} in the input array \texttt{a}.  In this example, the
access pattern is simple and easily understood. More complex
stencil computations are more prone to errors from simple lexical
mistakes. For example, the following is a small snippet from a
Navier-Stokes fluid simulator (based on~\citet{griebel1997numerical})
 in which two arrays are read with
different data access patterns, across two dimensions.
%%
\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-          &
  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
  /(4.0*delx)

duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+   &
  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-      &
  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-     &
  u(i,j))) / (4.0*dely)

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely

f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
\end{minted}
%%
The access pattern is much harder to understand than the
one-dimensional Laplace.  The miasma of indexing expressions of the
form \texttt{var(i$\pm$a, j$\pm$b)} is not only hard to read, but is
prone to simple textual input mistakes, \eg{}, swapping \texttt{-} and
\texttt{+}, missing an indexing term, or transforming the wrong
variable \eg{} \texttt{(i+1,j)} instead of \texttt{(i,j+1)}.

In practice, the typical development procedure for complex
stencil computations involves some ad hoc testing to ensure that no
transcription mistakes have been made (\eg{}, by visual
inspections, or comparison against an manufactured or 
analytical solutions~\cite{farrell2010automated}). Typically such
testing is discarded once the code is seen to be correct.

This is not the only information that is often discarded. The
``shape'' of the indexing pattern is usually the result of choices made in the
numerical-analysis procedure to convert some
continuous equations into a discrete approximation. Rarely are these
decisions captured in the source code, yet the shape of access is
usually uniform and is has a perspicuous and concise description in
numerical analysis literature \eg{}, ``\emph{centered in space, to a
  depth of 1}'' referring to indexing terms \texttt{a(i)},
\texttt{a(i-1)} and \texttt{a(i+1)}.

To support the development of correct stencil computations, we propose
a simple, abstract specification language for the data access pattern
of stencils. The shape of the
Laplace example stencil is specified in our language as:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
That is, \texttt{a} is accessed with a symmetrical pattern
(``centered'') to a depth of one in each direction in its first
dimension.  The Navier-Stokes example has a shape
specified as:
%
\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
!= stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u

!= stencil forward(depth=1,dim=1)                          * backward(depth=1,dim=2) :: v
\end{minted}
%
The specification requires that, over the whole fragment, \texttt{u}
is accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v} is
accessed in a neighbourhood bounded forwards to depth of $1$ in the
first dimension and backward to a depth of $1$ in the second
dimension. The specification is relatively small
and much more abstract compared to the source code.

We provide a verification tool which can check the spatial correctness
of stencil computations against a specification. Furthermore, the tool
can automatically infer specifications and insert these as comments
into the source code. The main features and benefits of our approach
are as follows:
%
\begin{itemize}
%\item The main use case is that a programmer writes a specification
%  first (perhaps informed by their numerical analysis), then writes
%  the corresponding source code, and lastly has the tool check its
%  conformance to the specifications. This specify-and-check approach
%  removes the usually ad hoc testing effort, and provides the
%  opportunity for easy re-verification in the event of any refactoring
%  of software maintenance activities.
\item The primary use case is that a programmer writes a specification
  first before the stencil code. Our tool then checks conformance between
  the two. This specify-and-check approach reduces testing effort and
  future-proofs against bugs introduced during refactoring and maintenance.
%
%\item The specification provides a concise description which can be
%  preserved in the source code to provide documentation to future
%  developers, contrasting with ad hoc tests which are usually
%  discarded or specialised to the development environment. Separate from the
%  verification procedure, the infer-and-synthesise mode of the tool
%  provides a way to automatically generate additional useful
%  documentation for future users.

% Emphasis here is on documentation generation, so it makes little sense to
% compare this aspect of CamFort to ad hoc testing. Apples and oranges.
\item A specification concisely captures the access pattern of a
  stencil and turns it into documentation. The inference procedure
  efficiently produces specifications with no programmer intervention,
  automatically inserting specifications at appropriate places in the
  source code.
%
%\item Specifications are integrated into the code as comments for easy
%  integration with existing tools, as well as documentation systems (\eg{}, we
%  support Doxygen and FORD syntax).
%
\item Our specification format is deliberately abstract, with a small number of
  combinators that \emph{do not involve any indexing expressions}, \eg{}
  \texttt{a(i+1,j-1)}. This contrasts with other specification
  approaches, \eg{}, deductive verification tools such as ACSL, where
  specifications must also be in terms of array-indexing expressions. Therefore, any low-level mistakes
  that could be made whilst programming complex indexing code could also be
  made when writing its specification. Our specifications are more
  abstract and lightweight
  with the aim of aiding adoption by scientists and preventing indexing errors.
\end{itemize}

Our verification tool does not target a
class of bugs that can be detected automatically (\emph{push-button
  verification}). Instead, stencil computation bugs must be
identified relative to a specification of the intended access pattern.

In this paper, we make the following contributions:
%
\begin{itemize}
\item We introduce a specification language for stencils
  that captures common forms of data access patterns
  (\cref{sec:lang});

\item We detail checking and inference algorithms for our
  specifications (\cref{sec:analysis}), derived from a
  denotational model for the specification language
  (\cref{sec:semantics});

\item We provide an implementation of our approach as an extension to
  CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;

\item We report on a quantitative study of stencil computations on a
  corpus of numerical Fortran program
  (\cref{sec:evaluation}), totalling one million lines of code. Our tool
  identifies and infers specifications for roughly 60,000 stencil computations in the
  corpus. Approximately 5,000 of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors. This validates
  our hypothesis that the majority of stencil computations have a regular
  shape, and validates the design of our language in its capability to capture
  many core patterns.

\item We give a verification case study of a
  stencil computation commonly used in scientific computing,
  simulating programming errors (\Cref{sec:jacobi}). Our approach detects all
  possible simulated programming errors for this problem.
\end{itemize}
%

\newcommand{\domainVal}{\mathbb{Z}_\infty}
\section{Stencil specifications}\label{sec:lang}

Our specification language is based on the hypothesis
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
\emph{five-point stencil} on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in \cref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

We give various definitions used in the paper then
\cref{subsec:syntax} outlines the syntax of stencil specifications
with an informal introduction. \cref{sec:semantics}
gives a semantic model via a set-based interpretation which makes
concrete the meaning of specifications. \cref{sec:eqs} then provides
a syntactic equational theory for specifications which is
sound with respect to the model.
%sub-specification for which the
%equality and sub-specification relation are proven sound.

\paragraph{Notation}
%\label{sec:notation}

\renewcommand*{\arraystretch}{0.8}
For the target language, $e$ ranges over expressions (may be impure) and $v$
over its (imperative) variables. %By
%``variable'', we mean in the imperative sense (named binders to memory cells).

\begin{definition}[Induction variable]
  An integer variable is an \emph{induction variable} if it is the control
  variable of a ``\texttt{for}'' loop incremented by $1$ per iteration. A variable
  is interpreted as an induction variable only within the scope of the loop
  body. Throughout, $i, j, k$ range over induction variables.
%TODO: consider mentioning derived, non-based induction variables
% as we did before.
\end{definition}

\begin{definition}[Array subscripts and indices]
  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
  which reads from an array $a$ at an \emph{index} specified by a
  comma-separated sequence of integer expressions denoted $\bar{e}$ or
  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$
  is called \emph{relative} if the expression involves an induction
  variable.
  An \emph{absolute index} is a constant integer expression relative to the
  enclosing loop.
\end{definition}

\begin{definition}[Neighbourhood index]
\label{def:neighbour}
  For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
  neighbourhood index is a constant translation of an induction variable. (The
  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
  relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
\end{definition}

\begin{definition}[Stencil computations]
\label{def:stencil}
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
  collection of arrays of arbitrary (possibly differing)
  dimensionalities.
  A \emph{stencil computation} comprises an iteration by a set $I$ of at
  most $n$ induction variables
  over a subset of the index space of $a$. Each iteration
  determines the elements of $a$ by an
  assignment $a(\bar{e}) = e_r$ where each $e \in
  \bar{e}$ is either a neighbourhood or constant index.
  For all $b \in \bar{b}$ and array subscripts $b(\bar{e}')$
  that flow to $e_r$, each $e' \in \bar{e}'$ is
  either a neighbourhood or absolute index.

  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
\end{definition}
%
% specifying the shape of indexing expressions for
%arrays $\bar{v}$ that are simultaneously permitted and expected within
%the code to which the specification is given.
%
%\marginpar{\dnote{Is it worth having this informal consistency here?}}

For now we informally describe what it means for a stencil kernel to
be consistent with a specification. This will be formalised in
\Cref{sec:semantics} and algorithmically described in
\Cref{sec:analysis}.  Specifications are associated to a stencil
kernel and are declared for one or more array variables $\bar{v}$
within that kernel.  A stencil specification defines a \emph{region}
associated to an array variable $a$ against which a stencil kernel
should be \emph{consistent}. The region defined by a specification is
an $n$-dimensional rectilinear shape in a discrete space crossing the
origin (which represents a collection of indices which are just 
induction variables). Consistency is then a
two way requirement. All array subscripts on $a$ contributing to the
kernel should be consistent with the region. Conversely, all parts of
the region should have an index corresponding it.

\begin{figure}[t]
\vspace{-0.9em}
\begin{align*}
\def\arraystretch{1}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{N}_{>0}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \nonterm{rvar} \; \\
\multicolumn{2}{l}{\qquad \mid \; \stenRefl{\dimTy}} \\
\multicolumn{2}{l}{\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
\multicolumn{2}{l}{\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
\multicolumn{2}{l}{\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
\multicolumn{2}{l}{\qquad \mid \; \nonterm{region} \, \term{+}
  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
%\\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\vspace{-0.8em}
\end{figure}

\subsection{Specification syntax}
\label{subsec:syntax}

\Cref{fig:syntax} gives the syntax of stencil specifications, which is
detailed below. The entry point is the \nonterm{specification} production which
splits into either a \emph{region declaration} or a \emph{specification
declaration}.
Regions comprise \emph{region constants} which are
combined via region operators \term{+} and \term{*}.

%\paragraph{Region constants}

Region constants specify a finite contiguous region in a
single dimension relative to the origin and are either \term{pointed},
\term{forward}, \term{backward}, or \term{centered}. The region names are
inspired by the scientific terminology, \eg{} the standard explicit method for
approximating PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
scheme~\citep{dawson1991finite}.

Each region
constant has a dimension identifier $d$ given as a positive natural number.
Each constant except \term{pointed} has a depth
parameter $n$ given as a positive natural number; \term{pointed}
regions implicitly have a depth of $0$.

A \term{forward} region of depth $n$ specifies a contiguous
region in dimension $d$ starting at the origin. This corresponds
to specifying neighbourhood indices in dimension $d$ ranging from $i$ to $i + n$
for some induction variable $i$. Similarly, a
\term{backward} region of depth $n$ corresponds to contiguous indices
from $i$ to $i - n$ and \term{centered}
of depth $n$ from $i - n$ to $i + n$. A \term{pointed}
stencil specifies a neighbour index $i$. For example, the
following shows four specifications with four consistent stencil
kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
e(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
!= stencil backward(depth=2, dim=1) :: b
e(i, j) = b(i, j) + b(i-1, j) + b(i-2, j)
!= stencil centered(depth=1, dim=2) :: c
e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
!= stencil pointed(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
Not every dimension needs to be specified, \eg{},
specifications on lines $1$, $3$, and $7$ leave some dimensions unspecified.

The \term{forward}, \term{backward}, and \term{centered} regions may
all have an additional attribute \term{nonpointed} which marks absence
of the origin.  For example, the following is a
\term{nonpointed} \term{backward} stencil
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, nonpointed) :: a
b(i) = a(i-1) + 10*a(i-2)
\end{minted}

\paragraph{Combining regions}

% r+s means that indices must be consistent with either r or s
% r*s means that indices must be consistent with both r and s at the
% same time

% specifications correspond to a set of schemes

% r+s means an index must be generalised by a scheme in either r or s
%       and every scheme in r must have an index which specialises it
%       and every scheme in s must have an index which specialises it

% r*s means an index must be generalised by a scheme in r and a scheme in s
%       and every pairs of schemes in r and s must have an index which
%           specialises both
%

The region operators \term{+} and \term{*} respectively combine
regions disjunctively and conjunctively.

The conjunction of two regions $r \term{*} s$ means that any indices
in the specified code must be consistent with both $r$ and $s$
simultaneously. Dually, for every pair of components in the regions
$r$ and $s$ there must be an index consistent with both.
For example, the following \emph{nine-point stencil}
has a specification given by the product of two \texttt{centered}
regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%%
This pattern is common in image convolution applications.
The specification ranges over the
values that flow to the array subscript on the left-hand side,
and so ranges over the intermediate assignments to \term{x},
\term{y}, and \term{z}. Each index in the code is consistent
with both specifications simultaneously, \eg{}, \texttt{a(i-1, j+1)}
is within the centered region in dimension $1$ and the centered region
in dimension $2$.

The product $r \term{*} s$ can also be thought of as a bounding box
over the two regions $r$ and $s$.

The disjunction of two regions $r \term{+} s$ means that any indices
in the specified code must be consistent with either of $r$ or $s$.
Dually, every part of region $r$ must have a
corresponding index and similarly for $s$ independently. For
example, the following gives the specification of a five-point
stencil which is the sum of two compound \texttt{pointed} and
\texttt{centered} regions in each dimension:
%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1)*pointed(dim=2)    + centered(depth=1, dim=2)*pointed(dim=1) :: a
b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
                   + a(i,j-1) + a(i,j+1)
\end{minted}
%%
Here the left-hand side of \texttt{+} says that when the second dimension
(induction variable $j$) is fixed at the origin, the first dimension
(induction variable $i$) accesses the immediate vicinity of the origin
(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
This reflects the symmetry under rotation of the five-point stencil.

\paragraph{Region declarations and variables}

Region specifications can be assigned to region variables via
region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common stencil patterns, such as the five-point
pattern, as the region can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in \Cref{fig:syntax}).

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

In some cases, it is useful to give a lower and/or upper bound for a
stencil. This can be done using either the \term{atMost} or
\term{atLeast} modifiers. This is particularly useful in situations
where there is a non-contiguous stencil pattern, which cannot be expressed
precisely in our syntax. For example:
%
\begin{minted}{fortran}
!= stencil atLeast, pointed(dim=1)         :: a
!= stencil atMost, forward(depth=4, dim=1) :: a
b(i) = a(i) + a(i+4)
\end{minted}

\section{A model for specifications}
\label{sec:semantics}

We define a simple set-theoretic denotational model of the semantics
of our specification language. This has a number of purposes:
(1) it serves to explain the meaning of our specifications
(2) it is used in the inference and checking algorithms
(\Cref{sec:analysis}); (3) justifies an equational theory for
specifications in the next section; and (4) can be used to guide
correct implementations. The model domain is over sets of vectors
which we call \emph{index schemes}.

\begin{definition}
  An \emph{index scheme} is a vector of size $n$ (an $n$-vector) with
  values drawn from $\domainVal = \mathbb{Z} \cup \{\infty\}$.  Values
  in $\mathbb{Z}$ represent the offsets of neighbourhood indices
  (\ie{}, the constant integer $c$ in an indexing expression of the
  form $i \pm c$, \Cref{def:neighbour}) and $\infty$ is a wild-card
  representing both constant indices and unspecified indices.
  For an index scheme $u$, the $i^{th}$ element is denoted $u_i$. We
  write $(\domainVal)^n$ for the set of all $n$-dimensional schemes.
\end{definition}

Index schemes abstract array subscripts.
Our model consists of index schemes explaining the range of possible
indexing behaviours which are consistent with the specification.
The relationship between source language
and index schemes
 is defined by \textsf{schematic} function,
 which forms part of checking and inference (\Cref{sec:analysis}).

>>>>

\begin{equation*}
\texttt{a} \mapsto \{[0,0]\times[0,0], \;\; [-1,0]\times[0,0], \;\;
[0,0]\times[-1,0], \;\; [0,0]\times[1,0], \;\; [0,1]\times[0,0]\}
\end{equation*}
%
We outline here the procedures for checking conformance
of source code against specifications (\Cref{subsec:checking})
and for inferring specifications from code (\Cref{subsec:inference}).
Both rely on a program analysis that converts array subscripts
 into sets of index schemes. We outline this analysis
first (\cref{subsec:analysis}). Note that the analysis
can be made arbitrarily more complex and wide-ranging independent
of the checking and inference procedures. At the moment, the analysis
is largely \emph{syntactic}, with only a small amount of
semantic interpretation of the code.
